\myTitle

### 2015-01-10 tex/main-body-text.tex
Got behind in commits, added script to simulate data, run flockture and structure and then do some comps

\section*{Comparison of {\sc structure} and {\sc flock} results}
Data sets were simulated (code provided on GitHub repository 
https://github.com/eriqande/flock-comment)
and evaluated in the programs {\sc flock} and {\sc structure} 
for the known value ($K=5$). Each dataset was composed of five populations of 
sizes 250, 250, 400, 550, and 550 comprised of individuals 
with 15 loci and an average pairwise $F_{ST} = 0.03, 0.02, 0.01, and 0.005$ among
populations. {\sc flock} was run nine times at $K=5$ with the default setting for all 
run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
LLOD threshold = 0). The default model in {\sc structure}, the no-admixture and correlated
allele frequency (NA-C) model, the no-admixture and non-correlated allle frequency (NA-NC) model, 
and the no-admixture and non-correlated allele frequency (NA-NC) model 
were run nine times with a 5,000 sweep burn-in followed by a 20,000 sweep sample 
from the posterior. All runs were performed on a JNCS D685 with an i7 3.40GHz processor and 8GB
of RAM running 64-bit Windows 7.

For each set of runs, {\sc flock} reports the 
normalized likelihood values (the exponentiated log-likelihood value for an individual's 
membership to each of $K$ clusters, scaled to sum to one) for a run belonging to the 
longest plateau - refered to as the 'best run' by \citep{Duc&Tur2012} (see below).
These values can be compared to the $q_i$ values
returned by {\sc structure}, which, in the no-admixture case, are posterior probabilities of cluster membership.
The program {\sc clumpp} \citep{Jak&Ros2007} was used to relabel clusters among different runs,
while the program {\sc district} \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.

\citet{Duc&Tur2012} describe a method for using {\sc flock} to estimate, $K$, the number of
subpopulations. The user must specify both the number of runs (different initial allocations of individuals 
to each cluster) as well as the number of iterations. (In each iteration every individual has the opportunity to be reallocated 
to a different cluster.) At the conclusion of each run, the log-likelihood value for membership to all $K$ clusters is computed
and the log likelihood difference (LLOD), the difference between the cluster with the highest likelihood 
and second highest likelihood, is calculated for each individual. {\sc flock} identifies runs that 
have converged to the same solution by detecting that they have the same mean LLOD score, and, in 
{\sc flock} parlance, a series of runs that have converged to the same solution are termed a ``plateau.'' 
The plateau record is a list of how many runs converged to the same allocation
of individuals. 
\citet{Duc&Tur2012} developed two stopping rules that they use to estimate $K$. 
\citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau of
$\geq 6$ runs is reached or four consecutive $K$ values yield no plateaus. If the first stopping condition is satisfied,
the $K$ for which there is a single plateau of $\geq 6$ runs is the point estimate for $K$. Alternatively, 
if multiple plateaus are observed with at least one plateau of $\geq 6$ runs, the largest value of $K$ that satisfies 
this conditon should be taken as the lower bound estimate of $K$. If no plateaus of $\geq 6$ runs are observed, $K$
is declared {\em undecided}.

%This section may get cut depending on how much we want to talk about estimating K.
It is now common practice to use the approximate marginal likelihood ($\ln P(D)$) from {\sc structure}
or a derived quantity such as $\Delta K$ \citep{Evannoetal2005} to try to estimate $K$.
While we take the stance that estimates of $K$ made with {\em any} unsupervised clustering algorithm
from data on real (non-idealized) populations should always be interpreted and used
cautiously, it is nonetheless instructive to compare \citet{Duc&Tur2012}'s method for
estimating $K$ to results obtained using $\ln P(D)$ and $\Delta K$, and we do so, below.

\subsection*{Results} 
Run times for both programs were relatively short, albeit in the case of {\sc flock} 
non-trivial. {\sc flock} averaged 32.8 minutes
for each run of the first simulated dataset, while each group of 
9 iterations for {\sc structure} averaged 14.7, 15.3 and 18.7 minutes
for the NA-NC, NA-C, and A-C models respectively.

{\sc flock} and {\sc structure} provide nearly identical results in 
terms of how individuals are allocated among runs for 
all simulations (Fig.~\ref{fig:Sims}). As $F_{ST}$ decreases 
between populations the solutions that {\sc flock} found were more variable.
Of the nine runs of {\sc flock} for the first simulated dataset, between 0 and 22
differences between the final arrangement of indivuals within the ($K=5$)
clusters. Of the nine runs, four pairs of runs found the 
same solution (runs 1 \& 2, 5 \& 6, 3 \& 7, and 4\&9). Run 8, which 
had no plateaus, was the most different with an average of 19.86 differences.
For the third simulated dataset, the number 
of differences ranged from 482 to 911.  {\sc structure} showed less variation for 
each set of simulations. Solutions for the first simulated dataset differed by the
placement of between 0 and 4 individuals and between 5 and 17 for the third 
dataset.
%Create a figure comparing all of the values. All the text is somewhat hard to read

Plateau lengths greater than 0 were only obsered for the first simulated dataset. For each
of the nine runs of {\sc flock} the plateau lengths varied from 0 to 4. In no analysis 
did we observe a plateau lenght greater than 6. Despite the similar allocation of individuals into $K$ clusters, the 
plateau analysis to estimate $K$ in {\sc flock} did not support $K = 5$.

but none greater than 5 while no plateaus were observed for either second or 
third simulated datasets.

\begin{figure*}
\begin{center}
    \includegraphics[width=\textwidth]{images/Figures-Pat/Fig_AllSim.pdf}   % Use this if we want it color (I'll make greyscale later)  
    \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values 
of {\sc flock} made with the program {\sc distruct}. Each horizontal plot represents one of nine individual runs of the program {\sc structure}, or the `best run' from a series of nine runs in {\sc flock} at $K=5$. Within each plot 
$q_i$ values from {\sc structure} and the normalized likelihood values 
from {\sc flock}  are represented by a 
%gray-scale
vertical bar. Each dataset is comprised of five populations of sizes 550, 550, 400, 250, and 250 individuals. The 
average pairwise $F_{ST}$ among populations was 0.03, 0.02, 0.01, and 0.005 for simulations 1, 2, 3, and 4 respecitively. 
}
    \label{fig:Sims}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
    \includegraphics[width=0.5\textwidth]{images/Figures-Pat/Loss.pdf}
    \caption{Comparision of zero-One loss for the admixture and correlated allele frequency (A-C) model, 
the no-admixture and correlated allele frequency (NA-C) model, the no-admixture and non-correlated 
allle frequency (NA-NC) model, the no-admixture and non-correlated allele frequency (NA-NC) 
model in {\sc structure} and the {\sc flock} model for each of the four simulated datasets.}
    \label{fig:Loss}
\end{center}
\end{figure*}

>>>>

\section*{Comparison of {\sc structure} and {\sc flock}}
The implementation of {\sc flock} as a VBA procedure (macro) in Microsoft Excel precludes
a rigorous evaluation of its behavior due to the inability to efficiently batch process. Consequenty, 
we adapted the source code from {\sc structure} as describe above to create {\sc flockture}
(code provided on GitHub repository https://github.com/eriqande/flockture) in order to accurately
assess the strengths and weaknesses of the {\sc flock} algorithm. First, using simulated 
data we demonstrated that {\sc flock} and {\sc flockture} are essentially equivalent programs.
Second, we simulated and analyzed datasets with {\sc flockture} and three models in {\sc structure} 
(the admixture and correlated allele frequency (A-C) model, 
the no-admixture and correlated allele frequency (NA-C) model, 
and the no-admixture and non-correlated allele frequency (NA-NC) model) to help better understand
under what conditions, if any, running {\sc flock} may be preferable to {\sc structure} .

For each set of runs, {\sc flock} reports the 
normalized likelihood values (the exponentiated log-likelihood value for an individual's 
membership to each of $K$ clusters, scaled to sum to one) for a run belonging to the 
longest plateau - refered to as the 'best run' by \citep{Duc&Tur2012} (see below).
These values can be compared to the $q_i$ values
returned by {\sc structure}, which, in the no-admixture case, are posterior 
probabilities of cluster membership.\citet{Duc&Tur2012} describe a method 
for using {\sc flock} to estimate, $K$, the number of
subpopulations. The user must specify both the number of runs (different initial allocations of individuals 
to each cluster) as well as the number of iterations. (In each iteration every individual 
has the opportunity to be reallocated 
to a different cluster.) At the conclusion of each run, the log-likelihood value for 
membership to all $K$ clusters is computed
and the log likelihood difference (LLOD), the difference between the cluster with the highest likelihood 
and second highest likelihood, is calculated for each individual. {\sc flock} identifies runs that 
have converged to the same solution by detecting that they have the same mean LLOD score, and, in 
{\sc flock} parlance, a series of runs that have converged to the same solution are termed a ``plateau.'' 
The plateau record is a list of how many runs converged to the same allocation
of individuals. \citet{Duc&Tur2012} developed two stopping rules that they use to estimate $K$. 
\citet{Duc&Tur2012} advocate running  {\sc flock} until either a single plateau of
$\geq 6$ runs is reached or four consecutive $K$ values yield no plateaus. If the first stopping condition is satisfied,
the $K$ for which there is a single plateau of $\geq 6$ runs is the point estimate for $K$. Alternatively, 
if multiple plateaus are observed with at least one plateau of $\geq 6$ runs, the largest value of $K$ that satisfies 
this conditon should be taken as the lower bound estimate of $K$. If no plateaus of $\geq 6$ runs are observed, $K$
is declared {\em undecided}.

To compare {\sc flock} and {\sc flockture} 4 datasets composed of 5 populations
were simulated using a coalescent based program (code provided on GitHub repository 
https://github.com/eriqande/flock-comment/simdata). Each of the 5 populations
were comprised of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes
and an average pairwise $F_{ST}$ of 0.03, 0.02, 0.01, and 0.005 among
populations. Both programs were run nine times at $K=5$ with the default setting for all 
run parameters (initial partition = random, number of iterations = 20, number of runs = 50, 
LLOD threshold = 0). We judged {\sc flockture} as an accurate representation of {\sc flock} based on the
following criteria: (1) Clustering of individuals into corrected population as defined by a zero-one loss function, 
(2) individual $q_i$ values, and (3) plateau records. All runs were performed on a JNCS D685 with an i7 3.40GHz 
processor and 8GB of RAM running 64-bit Windows 7.

Once we were confident that no major differences in inference existed between  {\sc flock} and {\sc flockture}
we used  {\sc flockture} to analyze a many more datasets that encompassed a broader 
range of population differentiation. Fourteen datasets were simulated as above (five populations 
comprised of 250, 250, 400, 550, and 550 individuals with 15 loci genotypes)
with average pairwise $F_{ST}$ values ranging between 0.11 and 0.005. 
 {\sc flockture} was run with the default run parameters and each run of 
{\sc structure} consisted of a 5,000 sweeep burn-in followed by a 20,000 sweep sample
from the posterior distribution. All runs were performed on a Macbook with a 2.3 GHz i7 processor
and 16GB of RAM running OS X ver. 10.7.5. The program {\sc clumpp} \citep{Jak&Ros2007} 
was used to relabel clusters among different runs,
while the program {\sc district} \citep{Rosenberg2004} was used to plot individual \textit{$q_i$} values.

\subsection*{{\sc flock} and {\sc flockture}} 
{\sc flock} and {\sc flockture} provided similar clustering of individuals among all four 
simulated datasets. The zero-one loss across all datasets had eqivalent results Fig.~\ref{fig:FvFloss}.
When genetic differentiation is high both programs correctly apportion the majority of individuals and 
as genetic differentiation degrades among populations both programs show corresponding declines in 
assignment ability. Neither the mean nor range of loss between {\sc flock} and {\sc flockture} differs
substantially among the simulated datasets. These assignments were dependent on the normalized likelihood
values ($q_i$), which while they were comparable may be slightly biased toward the bounds (0,1)
for {\sc flock}. In an attempt to understand the deviations being observed we took a subset of the 
first dataset (29 individuals from population 1 and 21 individuals from population 2) and initiated 
{\sc flock} with all individuals coming from population 1 starting in reference population 1 and all 
individuals from population 2 starting in reference population 2. {\sc flock} was run a second time
with the same starting clusters, but with the reference population labels switched. While the final allocation
of individuals matched the mean LLOD was slightly different (5.045009 vs. 5.028084). As a result, comparisons
between {\sc flock} and {\sc flockture} were limited to random initial starting points and should not 
be identical, but should have displayed equivalent behavior. Plateaus were only observed for the dataset 
with the highest pairwise $F_{ST}$ value $(0.03)$. Plateau records were similar, although 
the length of each plateau record was slightly longer for {\sc flockture} and on average had more runs converging on
the largest plateau (Supp. Table 1). Multiple runs of {\sc flockture} demonstrated that the differences are well within the
variability that we would expect from randomizing the initial grouping of individuals to 
initiate the algorithms.

Run times for both programs were relatively short, albeit in the case of {\sc flock} 
non-trivial. {\sc flock} averaged 1966.67 ($\pm 64.32$) seconds
over all runs, while {\sc flockture} averaged 9.83 ($\pm 0.21$) seconds. Run times showed a 
general increase as population structure became more subtle, however {\sc flock} did not 
demonstrate a linear increase. % this is surprising to me as it is coin the same thing irrespective of FST
% What I assume is happening is that as FST increases you have a lot more individuals being shuffled around
% at later iterations leading to increased process times. Seems reasonable.

\begin{figure*}
\centering
\begin{minipage}{.45\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{images/Figures-Pat/FlockvFlocktureLoss.pdf}
  \captionof{figure}{Zero-one loss function for {\sc flock} and {\sc flockture} for 9 replicate runs of simulated datasets comprised 
of five populations with varying levels of population differentiation:  $F_{ST} = $ (a) 0.03, (b) 0.02, (c) 0.01, and (d) 0.005.}
  \label{fig:FvFloss}
\end{minipage}%
\hfill
\begin{minipage}{.45\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{images/Figures-Pat/FlockvFlockture_qi.pdf}
  \captionof{figure}{Normalized likelihood values for {\sc flock} and {\sc flockture} for a single run of dataset 1 
with average pairwise $F_{ST} = $ 0.03. }
  \label{fig:FvFqi}
\end{minipage}
\end{figure*}

\subsection*{{\sc flockture} and {\sc structure}} 
As expected {\sc flockture} and {\sc structure} both perform extremely well on datasets with
well differentiated populations. Both programs have less than 1\% assignment error for most runs 
at values of $F_{ST}  \geq$ 0.0603 (Fig.~\ref{fig:FvSloss}). Infrequently, the 
NA-NC and NA-C models in {\sc structure} grouped two populations as one and 
assign individuals from another to two separate populations (Fig.~\ref{fig:FvSdistruct}  Simulation 1). This phenomena 
disappears as population differentiation declines. As the subtlety of genetic structure 
increased loss for both programs increases. 
Performance of the two programs as $F_{ST}$ approaches 0.0238 is relatively similar
characterized by the A-C model in {\sc structure} having the largest loss, the NA-C and NA-NC models performing
relatively similar and {\sc flockture} fluctuating around the results of the other three models. The scale of these 
differences is extremely small - on the order of 10 individuals being misassigned. As genetic 
differentiation continues to decline {\sc flockture}  showed consistently better performance 
at levels of $F_{ST}$ between 0.0238 to 0.0229.  Again, these differences are on the scale of roughly 20 individuals
or 1\%. In simulation 12 we observed the variation in {\sc flockture} results increase drastically (Fig.~\ref{fig:FvSloss}),
and for all other simulated datasets {\sc structure} had a lower assignment error.

The number of plateaus {\sc flockture} encountered and the number of runs that ended on the same plateau quickly 
diminished as the amount of differentiation among populations decreased. For simulations 1-4 a single plateau
greater than 6 was observed in every run. For simulation 5 only a single run yielded a plateau record 
with a single plateau greater than six. For simulations 6 to 11 plateau records of various lengths were observed
but all plateaus were less than 6. No plateaus were observed for datasets 12 through 20. Despite the lack of 
plateaus observed for these datasets graphs of the $q_i$ values were suggestive of population structure within the 
data (Fig.~\ref{fig:FvSdistruct}  Simulation 13). {\sc district} plots of the normalized likelihood values from
{\sc flockture} were consistent and suggestive of population structure within the dataset until simulation 17
after which the majority of the 'best runs' were substantially different in how they clustered individuals.

 \begin{figure*}
\centering
  \includegraphics[width=.9\linewidth]{images/Figures-Pat/FvSloss1.pdf}
  \caption{Zero-one loss over 9 replicate runs of the admixture and correlated allele 
  frequency (A-C) model, 
the no-admixture and correlated allele frequency (NA-C) model, 
and the no-admixture and non-correlated allele frequency (NA-NC) model in {\sc structure} and {\sc flockture}. 
Average pairwise $F_{ST}$ 
values between populations is listed above each graph.}
  \label{fig:FvSloss}
\end{figure*} % These graphs need to be redone so that text is legible and corrected for the NA-C NOT the A-NC 
%model in structure. I probably had a transcription error in R. Check code to be sure.

 \begin{figure*}
\centering
  \includegraphics[width=.8\linewidth]{images/Figures-Pat/FlockturevStructureDistruct.pdf}
  \caption{Plots of $q_i$ values from {\sc structure} and the normalized likelihood values from
  {\sc flockture} made with the program {\sc distruct}. Each horizontal plot represents one of nine 
 runs of the programs  at $K$ = 5. Three models were run in structure: the admixture and correlated allele 
  frequency (A-C) model, 
the no-admixture and correlated allele frequency (NA-C) model, 
and the no-admixture and non-correlated allele frequency (NA-NC) model. Each simulation is comprised of
 five populations of sizes 250, 250, 400, 550, and 550 individuals with average pairwise $F_{ST}$ values
 indicated above each plot.}
  \label{fig:FvSdistruct}
\end{figure*}

% haven't touched these yet...
